---
title: "PM 566 Final Project"
author: "Dana Gonzalez"
---

**Welcome to Dana Gonzalez's Final Project site for PM 566: Introduction to Health Data Science**

This project is a brief analysis of the rises in incidence of acute hepatitis C virus (HCV) across key demographics (namely race/ethnicity, age, and sex) in the United States from 2000 to 2022.

Please refer to the following link to view the full report with my complete analysis: [Download Report](https://github.com/danagonz/PM566-Final-Project/blob/1e157f918f3d900b85751f65d0b6f1a3e897a4c6/PM-566-Final-Written-Report.pdf).

# The Evolving Landscape of Acute Hepatitis C in the United States: A Demographic Breakdown

### Introduction

Hepatitis C virus (HCV) infection is the most common blood-borne infection in the United States. National surveillance data has shown a significant rise in acute HCV cases (those in the first 6 months of infection) in the last decade, with case counts having nearly doubled. The majority of acute cases (\>50%) will develop into chronic hepatitis C, which can lead to complications like liver disease, liver cancer, liver failure, or even death (Centers for Disease Control and Prevention \[CDC\], 2024).

Due to the asymptomatic nature of acute HCV, the true burden of this disease is grossly underestimated. For this reason, it is critical that current at-risk populations are promptly identified to more effectively target public health interventions to connect at-risk and diseased individuals with appropriate preventative, diagnostic, and therapeutic care.

**For this reason, I will work to visualize trends in acute HCV incidence across the above-mentioned demographic groups for the period 2000-2022 and determine which group(s) within each demographic factor have seen the greatest rise in incidence for the period of interest.**

**I will evaluate rises in incidence by fitting linear models to plots of each group and examine each subgroup's slope-intercept equation. I will interpret the group(s) with the greatest positive slope as having the greatest rise in incidence for the period of interest.**

### Methods

All data for this project was retrieved from the Centers for Disease Control and Prevention's (CDC) National Center for HIV, Viral Hepatitis, STD, and Tuberculosis Prevention (NCHHSTP).

### Results

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Load Libraries
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(devtools)
library(MoMAColors)
library(plotly)
library(zoo)
library(tidyr)
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Read in Data
estimated <- read.table("~/Downloads/Reported_vs_Estimated.csv", header = TRUE, sep = ",")
race <- read.table("~/Downloads/Race_Data.csv", header = TRUE, sep = ",")
age <- read.table("~/Downloads/Age_Data.csv", header = TRUE, sep = ",")
sex <- read.table("~/Downloads/Sex_Data.csv", header = TRUE, sep = ",")
state <- read.table("~/Downloads/2022_State_Rates.csv", header = TRUE, sep = ",")
urbanicity <- read.table("~/Downloads/Urbanicity_Data.csv", header = TRUE, sep = ",")
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Load Libraries
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(devtools)
library(MoMAColors)
library(plotly)
library(zoo)
library(tidyr)
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Read in Data
estimated <- read.table("~/Downloads/Reported_vs_Estimated.csv", header = TRUE, sep = ",")
race <- read.table("~/Downloads/Race_Data.csv", header = TRUE, sep = ",")
age <- read.table("~/Downloads/Age_Data.csv", header = TRUE, sep = ",")
sex <- read.table("~/Downloads/Sex_Data.csv", header = TRUE, sep = ",")
state <- read.table("~/Downloads/2022_State_Rates.csv", header = TRUE, sep = ",")
urbanicity <- read.table("~/Downloads/Urbanicity_Data.csv", header = TRUE, sep = ",")
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Clean and Wrangle Data
## Race
race <- race |>
  select(-Indicator)
race <- race |>
  select(-Cases)
race <- race[!(race$Race.Ethnicity %in% c("Other", "Unknown")), ]
race <- race |>
  rename(
    year = Year,
    group = Race.Ethnicity,
    incidence = Rate.per.100000)
race$group <- as.factor(race$group)
race$year <- gsub("\\(COVID-19 Pandemic\\)", "", race$year)
race$year <- as.numeric(race$year)
race$incidence <- gsub("[^0-9.]", "", race$incidence)
race$incidence <- as.numeric(race$incidence)

## Age
age <- age |>
  select(-Indicator)
age <- age |>
  select(-Cases)
age <- age |>
  rename(
    year = Year,
    group = Age.Group,
    incidence = Rate.per.100000)
age$group <- as.factor(age$group)
age$year <- gsub("\\(COVID-19 Pandemic\\)", "", age$year)
age$year <- as.numeric(age$year)
age$incidence <- gsub("[^0-9.]", "", age$incidence)
age$incidence <- as.numeric(age$incidence)

## Sex
sex <- sex |>
  select(-Indicator)
sex <- sex |>
  select(-Cases)
sex <- sex |>
  rename(
    year = Year,
    group = Sex,
    incidence = Rate.per.100000)
sex$group <- as.factor(sex$group)
sex$year <- gsub("\\(COVID-19 Pandemic\\)", "", sex$year)
sex$year <- as.numeric(sex$year)
sex$incidence <- gsub("[^0-9.]", "", sex$incidence)
sex$incidence <- as.numeric(sex$incidence)

## State
state <- state |>
  select(-Range)
state <- state |>
  filter(State.or.jurisdiction != "District of Columbia")
state <- state |>
  rename(
    abbreviation = State,
    state = State.or.jurisdiction,
    incidence = Cases.100.000.population)
state$state <- as.factor(state$state)
state$incidence <- gsub("[^0-9.]", "", state$incidence)
state$incidence <- as.numeric(state$incidence)

## Urbanicity
urbanicity <- urbanicity |>
  rename(
    year = Year,
    area = Area,
    incidence = Incidence)
urbanicity$area <- as.factor(urbanicity$area)
urbanicity$year <- as.numeric(urbanicity$year)
urbanicity$incidence <- gsub("[^0-9.]", "", urbanicity$incidence)
urbanicity$incidence <- as.numeric(urbanicity$incidence)
```


#### **Yearly Incidence (per 100,000) by Race/Ethnicity (2000-2022)**

```{r, echo=FALSE, fig.width=9, fig.height=6}
# Interactive Lineplot of Yearly Incidence (per 100,000) by Race/Ethnicity (2000-2022)
plot_ly(race, 
        x = ~year, 
        y = ~incidence, 
        color = ~group, 
        type = "scatter", 
        mode = "lines", 
        line = list(width = 4),
        colors = c("#ff9796", "#db626a", "#a91741", "#670b34", "#44302a")) |>
  layout(
    xaxis = list(title = "Year"),
    yaxis = list(title = "Incidence Rate"))
```

#### **Yearly Incidence (per 100,000) by Age Group (2000-2022)**

```{r, echo=FALSE, fig.width=9, fig.height=6}
# Interactive Lineplot of Yearly Incidence (per 100,000) by Age Group (2000-2022)
plot_ly(age, 
        x = ~year, 
        y = ~incidence, 
        color = ~group, 
        type = "scatter", 
        mode = "lines", 
        line = list(width = 4),
        colors = c("#ff9796", "#db626a", "#a91741", "#670b34", "#625856","#44302a", "#090707")) |>
  layout(
    xaxis = list(title = "Year"),
    yaxis = list(title = "Incidence Rate"))
```

#### **Yearly Incidence (per 100,000) by Sex (2000-2022)**

```{r, echo=FALSE, fig.width=9, fig.height=6}
# Interactive Lineplot of Yearly Incidence (per 100,000) by Sex (2000-2022)
plot_ly(sex, 
        x = ~year, 
        y = ~incidence, 
        color = ~group, 
        type = "scatter", 
        mode = "lines", 
        line = list(width = 4),
        colors = c("#ff9796", "#a91741")) |>
  layout(
    xaxis = list(title = "Year"),
    yaxis = list(title = "Incidence Rate"))
```

#### **Incidence Rates by U.S. State (2022)**

**Note:** States with incidence rates of "-1" did not have data available for 2022. They were coded as such to allow for display in Figure 5.

```{r, echo=FALSE, fig.width= 9, fig.height=6, message=FALSE, warning=FALSE}
# Map of Incidence Rate by US State
state$incidence[is.na(state$incidence)] <- -1
shadeLimit <- max(state$incidence[state$incidence != -1], na.rm = TRUE)
state_map <- plot_geo(state, locationmode = 'USA-states') |> 
  add_trace(
    z = ~incidence, 
    text = ~state, 
    locations = ~abbreviation,
    color = ~incidence, 
    colorscale = list(
      list(0, "white"),
      list(0.01, "#ff9796"),
      list(1, "#a91741")),
    showscale = TRUE) |> 
  layout(
    geo = list(
      scope = 'usa', 
      projection = list(type = 'albers usa'),
      center = list(lon = -98.5, lat = 39.8),
      showland = TRUE,
      landcolor = "white",
      subunitwidth = 1,
      countrywidth = 2),
    colorbar = list(
      title = "Incidence (per 100,000)",
      limits = c(0, shadeLimit)))

state_map
```

### Conclusion and Summary